Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-14 agujournaltemplate.tex
Final version submitted to JGR Biogeosciences

    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were obtained from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}. In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
    For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a gradient forest analysis. Gradient forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community changes. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$; see Table \ref{tab:DataSources} for a full overview). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the gradient forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

@@

%!
\section*{Open Research Section}
All data used in this manuscript are publicly available. The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}. This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset \url{https://www.bco-dmo.org/dataset/3095}.
The climate variables, such as wind speed, precipitation and evaporation, were extracted from the ``ERA5 monthly averaged data on single levels from 1940 to present'' dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.
All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
%TODO add Zenodo link before submission

>>>>

    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were obtained from the Biological and Chemical Oceanography Data Management Office (BCO-DMO). 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size). For simplicity, we refer to this variable as ``wind speed'' throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems. The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}.

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}. In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}. All the codes we produced for data processing and data analysis are publicly available on GitHub \cite{benjamin_post_2025_16878343}.
    For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S. National Oceanographic Data Center \cite{johnson2006world}. The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis. Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}. For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a gradient forest analysis. Gradient forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community changes. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$; see Table \ref{tab:DataSources} for a full overview). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the gradient forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub at \url{https://github.com/ben1post/PhytoCariaco} \cite{benjamin_post_2025_16878343}.

@@

%!
\section*{Open Research Section}
All data used in this manuscript are publicly available. The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}. This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data at \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset at \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset at \url{https://www.bco-dmo.org/dataset/3095}.
The climate variables, such as wind speed, precipitation and evaporation, were extracted from the ``ERA5 monthly averaged data on single levels from 1940 to present'' dataset at \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.
All the code and data used in this analysis and to create the plots are publicly available on GitHub at \url{https://github.com/ben1post/PhytoCariaco} \cite{benjamin_post_2025_16878343}.
